Inferring causal connectivity from pairwise recordings and optogenetics

### 2018-10-23 manuscript/main.tex
version 1.0

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usetikzlibrary{arrows,shapes,calc}
\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\newcommand{\mean}[1]{\left<{#1}\right>}
\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}
\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Inferring causal connectivity from pairwise recordings and optogenetics}
%\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
\author{Mikkel Elle Lepper\o{}d, Tristan St$\ddot{\text{o}}$ber, Konrad Paul Kording}

\begin{document}

 \maketitle

\begin{abstract}
\noindent Neuroscientists often want to identify how neurons causally interact. Perturbations with optogenetics are a popular way of quantifying such interactions. However, even focused light will typically stimulate multiple cells, which produces a confound - we cannot know which of the stimulated neurons affected the activity of a given target neuron. Here we show how this confound produces large biases in estimates of connectivity, and how we can reduce these biases by a data analysis strategy that exploits the absolute refractory period of neurons. We use the instrumental variable (IV) framework from economics which estimates causal strengths by exploiting a variable that affects only one of the variables, here the presynaptic neuron. The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem. In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 0.6\% correct). Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
\end{abstract}

\section{Introduction}

\todo[inline]{Cite Konrad's new paper}
% neuroscience experiments are observational which implicates a certain set of statistical tools
% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
%a quest for causality and optogenetics is a main tool
% We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
%set up the drive to causality
We want to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, action, and cognition. For such an understanding it is not sufficient to know the correlations between variables or even be able to predict them. After all, there can be many ways how the same activities come about by distinct causal chains \citep{drton2011global, peters2017elements}. Identifiability of causal networks is complicated when it comes to brain data. Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. Thus, brain data which is almost always observational, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. While observing correlations within the system is usually relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard or, more commonly, impossible.  Getting at such an understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons. Only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}. Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

% we use observational studies. But causality is typically wrong. 
Moreover, we generally only record from a small subset of all neurons. The data we obtain from typical recordings, e.g. from electrophysiology  or calcium imaging, is very low dimensional relative to the dimensionality of the brain. Moreover, it is observational, which means that it does not result from randomized perturbations. In such cases, we can never know to which level the observed activity was caused by other observed activity, or by the activity of the unobserved neurons. Such unobserved activity is then called confounders. If mechanisms are estimated from observational data in the presence of confounders, we will generally make large errors and draw incorrect conclusions \citep{angrist2008mostly}. Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.

%set up confounding
Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.  A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM).  Multiple regression may be perceived as a solution to confounding problems as they support the concept of ``explaining away" \citep{stevenson2008inferring}. However, in current work we will exclude approaches that look at multiple neurons at the same time. This is because, ``explaining away" can only be a good strategy if most neurons are included in the recordings but this is rarely the case for most experimental settings, especially in the mammalian brain. We will therefore focus on neuron pairs where stimulus and the pre- and post-synaptic neurons are known.

To estimate connectivity it is first and foremost important that it stems from cause and effect e.g. B's action potential is causing an increase in C's membrane potential, therefore we use the term causal connectivity.  For a given example we want to estimate causal connectivity between two neurons, $ A $ and $ C $ (\cref{fig:intro}\labelcref{fig:intro:1}). Two neurons, $ A $ and $ B $, are driven by a common input $ D $ and because $ B $ and $ C $ are connected they are strongly correlated. Consequently, $ A $ and $ C $ are also correlated and a regression  $ C = \beta A $ will misleadingly conclude that there is a direct interaction if it was causally interpreted.  In this case we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. Na\"ive regressions in partially observed systems will generally not reveal causality.

To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard. In fact, a common definition of causality is in terms of what would happen if one would change the value of one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}. If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. However, this is experimentally challenging and yield low cell count because it either requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}. Because gold-standard perturbations are challenging, it would be highly desirable if causality could be obtained from optogenetic stimulation  \citep{boyden2005millisecond, zemelman2002selective}.

%optogenetics is totally nonlocal
Interpreting the results of optogenetic stimulation in terms of causal interactions is difficult. The expression of opsins are often restricted to a population of neurons, but regular optogenetic stimulation will still affect many neurons simultaneously. In \textit{in vivo} settings it is generally impossible to direct photons to only one neuron. Hence, the stimulus will produce a distributed pattern of activity. For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside. This distributed pattern of stimulation produces activity which then percolates through the network of neurons. Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.

%causal inference is a thing, if not done well you say stupid things
The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}. Within these fields, the problem of endogenous regressors is commonly addressed. We may thus look towards these fields for insights into how we may resolve the confounding problem induced by optogenetic stimulation.

%example
A commonly used approach towards causal inference in economics are instrumental variables. Let's say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. Here $ u $ are the factors other than education that contribute to yearly wages. One of the factors in $ u $ is a person's cognitive ability. However, a person's cognitive ability may also affect education and thus the regressor $ x $  is correlated with the error term $ u $. This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. Proximity is a good instrument since it can only affect wages through schooling, since we expect living in proximity of colleges and universities give higher probability to attend but not to affect other contributing factors to wages such as cognitive ability. Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$. This ratio corrects for the confounding factor.

%worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}. First, the instrument must be uncorrelated with the error term. Second, the instrument must be correlated with the regressor. Third, there must be no direct influence of the instrument on the outcome variable but only an influence through the regressor variable. The validity of these assumptions is central when using the IV.

% Refractory period as IV
For an instrument to be good, it needs to be unaffected by other variables. In the brain, almost everything is affected by the network state. However, certain variables can be more or less affected. For example, the overall activity of the network is through slow and strongly nonrandom dynamics. In contrast, the temporal pattern of when a neuron is in a refractory state may be random. First, if neurons are spiking according to conditional Poisson distributions, their exact timing will, conditioned on the network state, be random. While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is likely quite random.

%here we
Here we show that the instrumental variable (IV) technique can be employed if one seeks to estimate the causal connectivity between neuron pairs. We begin by showing how confounding factors are introduced by regular optogenetic stimulations. We then simulate this confounding effect in a simple network of three leaky integrate and fire (LIF) neurons. With this simple model we show that by using the refractory period as an IV we are able to distinguish between connected and unconnected neuron pairs. We compare these estimates with a naive although widely used cross-correlation histogram (CCH) method that fails to distinguish respective pairs. We then turn to a simulated network of randomly recurrent connections of excitatory and inhibitory LIF neurons with distributed synaptic weights. With this data at hand we first calculate the mean squared errors of the IV method and show that it is robust to different simulated network states. Finally, we compare the amount and size of false positive and false negative estimates and goodness of fit on synaptic weights with pairwise assessments using CCH and logistic regression.

\section{Results}
\subsection{Optogenetics is not local}
Optogenetics is generally seen as a perturbation method that by and large affects neurons in close proximity of the light source. However, it can be questioned if this is the correct way of conceptualizing the spatial effect of stimulation. The stimulation effects depend on multiple factors. Firstly, light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. Secondly, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. Lastly, physiological properties of the cells are important as light may e.g. only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. The induced effect of optogenetic stimulation as a function of distance should be given by a relation of the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.

To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}. This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset). The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. The density of photons thus decreases rapidly with distance.

The number of illuminated neurons at a given distance will be larger the further away they are from the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line). In fact, it will increase by approximately $ r^2 $ with distance. This derives from the same surface scaling as for the 3D shell. Thus the number of neurons that can be activated increases rapidly with distance.

To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. This can largely be characterized by the  distribution of membrane potentials across neurons. Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current. The peak amplitude of the photo current relates non-linearly to the light intensity and can be modelled by the Hill equation \citep{wang2007high}. Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even. Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current which actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.

\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line). The total photo current (sum of single neuron currents) thus increases by distance due to the nonlinear relation between photo intensity and photo current (P, blue line). \label{fig:concept}}
\end{figure}
\FloatBarrier
\subsection{Confounding as a problem for the estimation of causal effects}
When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity. To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C) \cref{fig:intro}\labelcref{fig:intro:1}. The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential. We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions.
Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.

After running the simulation, the peri-stimulus time histogram of the stimulated neurons show the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}\labelcref{fig:intro:2} (AA, BB).  Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:3} (AB). This further generates a strong correlation between A and C, confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}) shown in \cref{fig:intro}\labelcref{fig:intro:3}. Even though the correlation peak between B and C is larger than between A and C due to correlated spikes outside the periods with stimulation one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected. If stimulation affects multiple neurons simultaneously, there is a real confounding problem.

\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{raster}
\caption{} \label{fig:intro:2}
\end{subfigure}

\medskip
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf A,B,C network}. A sketch of the simple network containing three neurons %depicted in \labelcref{fig:intro:3} 
shows stimulation configuration with blue laser light and the connections with arrows (a). The neurons A and B are stimulated 1000 trials and the corresponding peristimulus time histogram are shown in \labelcref{fig:intro:2} upper panel with a raster plot in the lower panel. Cross correlation histograms (CCHs) are shown in \labelcref{fig:intro:3} where horizontal and vertical axes represents time lag in ms and counts of coincident spikes in bins of 1 ms. \label{fig:intro}}
\end{figure}
\FloatBarrier
\subsection{Instrumental variables to resolve confounding}
In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons we need something that can distinguish the influence of one stimulated neuron from the influence of another stimulated neuron. We would thus need something that affects the stimulation effect separately between neurons. Arguably, refractoriness is such a variable. If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. This gives us an interesting way of getting at causality, by comparing the network state between a time when a neuron is able to spike and a time where the neuron is unable to spike.

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{url}

\usetikzlibrary{arrows,shapes,calc}

\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\newcommand{\mean}[1]{\left<{#1}\right>}
\newcommand{\cor}[2]{\mathrm{cor}\left(#1, #2\right)}

\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}

\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Inferring causal connectivity from pairwise recordings and optogenetics}
%\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
\author[1,2]{Mikkel Elle Lepper\o{}d}
\author[2,3,4]{Tristan St$\ddot{\text{o}}$ber}
\author[1,2]{Torkel Hafting}
\author[2,5]{Marianne Fyhn}
\author[6]{Konrad Paul Kording}
\affil[1]{Institute of Basic Medical Sciences, University of Oslo, Oslo, Norway}
\affil[2]{Centre for Integrative Neuroplasticity, University of Oslo, Oslo, Norway}
\affil[3]{Department of Informatics, University of Oslo, Oslo, Norway}
\affil[4]{Department of Computational Physiology, Simula Research Laboratory, Lysaker, Norway}
\affil[5]{Department of Bioscience, University of Oslo, Oslo, Norway}
\affil[6]{Department of Neuroscience, University of Pennsylvania, Pennsylvania, USA}
\begin{document}

 \maketitle
% suggestions for referee?
\begin{abstract} % max words 150
\noindent To study how the brain works, it is essential to identify causal interactions between neurons, usually requiring perturbations. However, when using optogenetics we typically perturb multiple cells, producing a confound - any of the stimulated neurons can have affected the target neuron. Here we show how this produces large biases, and how they can be reduced using the instrumental variable (IV) framework from econometrics. The interaction between stimulation and the absolute refractory period produces a weak, approximately random signal which can be exploited to estimate causality. When simulating integrate-and-fire neurons, we find that IV estimates connectivity better than na\"ive techniques ($R^2=0.845$ vs $R^2=0.048$). The difference is important as the estimates differ when applied to experimental data from stimulated neurons with recorded spiking activity. Our IV approach is easy to implement in the context of current experimental paradigms.

\end{abstract}

\section{Introduction}
% The brain is infinitely confounded, thus correlations are not sufficient to draw conclusions about mechanisms and this is a hard problem in brain data.
A goal of neuroscience is to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, cognition, and action. 
Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. 
Because of this, it is not sufficient to know the correlations between variables or even be able to predict them.
While observing correlations within the system is relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard.
After all, there can be many ways that the same activities emerge from distinct causal chains \citep{drton2011global, peters2017elements}. 
Reaching a mechanistic level of understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons.
Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

% We use observational studies, where identifiability is questionable and causality is thus typically wrong. 
Typically, we only record from a small subset of all neurons. 
The data we obtain from such recordings, e.g. from electrophysiology or calcium imaging, is observational, which means that it does not result from randomized perturbations. 
In such cases, we can never know to which level the observed activity was caused by other observed activity, or by unobserved activity. 
The unobserved neurons are called confounders.
If mechanisms are estimated from observational data in the presence of confounders, the consequence may be large errors and incorrect conclusions \citep{angrist2008mostly}. 
Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.

% Set up confounding
Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.  
To estimate connectivity it is first and foremost important that it stems from cause and effect, therefore we use the term causal connectivity. 
Consider an example where we want to estimate causal connectivity between two observed, but unconnected neurons, $ A $ and $ C $ (\cref{fig:intro}(a)). 
A third unobserved neuron $ B $, is driven together with $ A $ by a common input and they are thus strongly correlated.
Furthermore, $ B $ and $ C $ are connected, thus the input drives $ C $ through $ B $.
Consequently, $ A $ and $ C $ are also correlated and the regression $ C = \beta A $ will misleadingly conclude a direct interaction when causally interpreted. 
In this case, we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. 
Na\"ive regressions in partially observed systems will generally not reveal causality.

% Why do we not use GLM
A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM). 
Multiple regression may be a solution to confounding problems as they support ``explaining away" background activity \citep{stevenson2008inferring}.
However, this is only a good strategy if most neurons are included in the recordings.
Furthermore, only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
This is rarely the case in experimental settings, especially in the mammalian brain.
Thus, brain data, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. 
We thus aim to develop methods that can identify causal connectivity between neuron pairs.

% Stimulation is an intervention which is needed to estimate causal interactions
To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard. 
In fact, a common definition of causality is in terms of the effect of changing one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}. 
If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. 
However, this is experimentally challenging and yield low cell count because it requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}. 
Because gold-standard perturbations are challenging, it is necessary and highly desirable if causality could be obtained from optogenetic stimulation in combination with neural recordings of large populations of neurons \citep{boyden2005millisecond, zemelman2002selective}.

% Optogenetic stimulation is usually totally non-local, producing confounders
Interpreting the results from optogenetic stimulation in terms of causal interactions is difficult. 
In most experimental settings, optogenetic stimulation will affect many neurons simultaneously. 
Hence, the stimulus will produce a distributed pattern of activity.
This distributed pattern of stimulation produces activity which then percolates through the network of neurons. 
Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.

% Causal inference is a thing, if not done well you say stupid things
For insights into how we may resolve the confounding problem induced by optogenetic stimulation, we may look to other fields that have addressed the problem of endogenous regressors. 
The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}.

% Example
A commonly used approach towards causal inference in economics are instrumental variables. 
Let's say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. 
Here $ u $ are the factors other than education that contribute to yearly wages. 
One of the factors in $ u $ is a person's cognitive ability which may also affect education and thus the regressor $ x $  is correlated with the error term $ u $. 
This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. 
In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. 
Proximity is a good instrument since it can only affect wages through schooling. 
We expect living in proximity of colleges and universities to give higher probability to attend without affecting other contributing factors to wages such as cognitive ability. 
Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$. 
This ratio corrects for the confounding factor.

% Worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}. 
First, the instrument must be decorrelated with the error term. 
Second, the instrument must be correlated with the regressor. 
Third, there must be no direct influence of the instrument on the outcome variable, but only an influence through the regressor variable. 
The validity of these assumptions is central when using the IV approach.

% Refractory period as an IV
For an instrument to be good, it needs to be unaffected by other variables. 
In the brain, almost everything is affected by the network state. 
However, certain variables can be more or less affected. 
For example, the overall activity of the network is due to slow and strongly nonrandom dynamics. 
In contrast, the temporal pattern of when a neuron is in a refractory state may be random. 
First, if neurons are spiking according to conditional Poisson distributions, their exact timing conditioned on the network state, will be random. 
While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is sufficiently random.

% Here we
Here we show that the IV technique can be employed if one seeks to estimate the causal connectivity between neuron pairs. 
We begin by showing how confounding factors are introduced by conventional optogenetic stimulations. 
We then simulate this confounding effect in a simple network of three leaky integrate and fire (LIF) neurons. 
With this simple model we show that by using the refractory period as an IV we are able to distinguish between connected and unconnected neuron pairs. 
We compare these estimates with a na\"ive, although widely used, cross-correlation histogram (CCH) method that fails to distinguish respective pairs. 
We then turn to a simulated network of randomly recurrent connections of excitatory and inhibitory LIF neurons with distributed synaptic weights. 
With this data at hand, we first calculate the mean squared errors of the IV method and show that it is robust to different simulated network states. 
We also compare the amount and size of false positive and false negative estimates and goodness of fit on synaptic weights with pairwise assessments using CCH and logistic regression.
Finally, we tested the methods on experimental data from extracellular recordings of single unit activity from optogenetically perturbed ensembles of neurons in the hippocampus.
The observed differences between the IV and the CCH estimate, underline the importance of considering potential confounding, when estimating connections based on neural activity.

\section{Results}
\subsection{Optogenetics is not local}
Optogenetic stimulation is generally seen as a perturbation that by and large affects neurons in proximity (< 1mm) of the light source with the effect decreasing with distance. 
However, this may be an incorrect way of conceptualizing the spatial effect of stimulation as it depends on multiple factors. 
Light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. 
Moreover, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. 
Finally, physiological properties of the cells are important as light may have a stronger effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. 
The induced effect of optogenetic stimulation as a function of distance should be given by a relation between the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and the distribution of induced photo current.

To estimate the light intensity, we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}. 
This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). 
This is explained by the surface of a 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset). 
The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. 
As a result, the density of photons decreases rapidly with distance.

The number of illuminated neurons at a given distance will increase with distance to the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line). 
In fact, it will increase by approximately $ r^2 $ with distance. 
This derives from the same surface scaling as for the 3D shell. 
Thus the number of neurons that can be activated increases rapidly with distance.

To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. 
This can largely be characterized by the  distribution of membrane potentials across neurons. 
Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. 
The expected response from a pulse of light that induces a charge $Q$ should be proportional to the number of affected neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). 
Given that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current. 
The peak amplitude of the photo current relates approximately logarithmically to the light intensity \citep{wang2007high}; see \cref{sec:method:opto}. 
Assuming that opsins are evenly distributed across neurons, the induced photo current will not be proportional with light intensity - it will fall slower. 
Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current. 
This product actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). 
Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.

\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. 
Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. 
If neurons are uniformly distributed, the number of affected neurons in a spherical slice increases by $ r^{2} $ (N, black line). 
The total photo current (sum over neurons of peak amplitude photo current in a spherical slice) thus increases with distance due to the nonlinear relation between light intensity and photo current (P, blue line) depicted as percentage of maximum.}
\label{fig:concept}
\end{figure}

\FloatBarrier
\subsection{Confounding as a problem for the estimation of causal effects}
When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity. 
To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C) \cref{fig:intro}(a). 
The neurons receive Poisson spike trains and have Gaussian white noise added to the membrane potential. 
Neurons were also interacting, where spikes of neuron B increase the probability of firing for neuron C, but there were no other interactions.
Finally, we allowed simulated optogenetic stimulation (current pulse) to affect neurons A and B (but not C). 
We thus have a simple system for exploring questions of causality.

After running the simulation, the peri-stimulus time histogram of the stimulated neurons (\cref{fig:intro}(b)) shows the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}(c) (AA, BB).
Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}(c) (AB). 
This further generates a strong correlation between A and C, confounding the system by rendering the cross-correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}). 
Even though the correlation peak between B and C is larger than between A and C, due to correlated spikes outside the periods with stimulation, one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected. 
If stimulation affects multiple neurons simultaneously, there is a real confounding problem.

\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{psth_triple}
\caption{} \label{fig:intro:2}
\end{subfigure}\medskip\\
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf Optogenetic stimulation induces spurious correlations}. 
A sketch of the simple network containing three neurons shows stimulation configuration with blue laser light and the connections with arrows (a). 
The neurons A and B are stimulated in 1000 trials and the corresponding peristimulus time-histogram are shown in (b) upper panel with a raster plot in the lower panel. 
Cross-correlation histograms (CCHs) are shown in (c) where horizontal and vertical axes represent time lag in ms and counts of coincident spikes in bins of 1 ms.}
\label{fig:intro}
\end{figure}
